Assessing the Population-level Conservation Effects of Marine Protected Areas

### 2021-05-02 documents/ovando-et-al-pop-effects-of-mpas-si.Rmd
pre-print cleanup

---
title: "Appendix Supporting Information 1"
subtitle: "Assessing the Population-level Conservation Effects of Marine Protected Areas"
output: 
  bookdown::pdf_document2:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: template.docx
  pdf_document: default
bibliography: references.yaml
csl: conservation-biology.csl
header-includes:
    - \usepackage{longtable}
    - \usepackage{booktabs}
    - \usepackage{eso-pic}
    - \usepackage{graphicx}
    - \usepackage[left]{lineno}
    - \usepackage{xcolor}
params:
  run_name: ["v6.0"]
  sim_years: [50]
  burn_years: [20]
  num_patches: [25]
always_allow_html: true
---

@@

@@

All code needed to reproduce our main results and manuscript can be found at https://github.com/DanOvando/population-effects-of-mpas. All analysis were performed in `r R.version.string`. Package versions are shown in Table.\@ref(tab:pkg-info).

@@

@@

## Recruitment

@@

## Dispersal

@@

## Fleet Dynamics

@@

## Spatial Fleet Distribution

@@

## MPA Design

@@

## Simulations

@@

### Filtering Simulations

@@

### Additional Simulation Results

@@

The full table of covariates and their posterior means and 89% credible interval can be found in Table.\@ref(tab:did-coefs). "var" refers to variable, "_2" indicates a squared term. tex is the mean cumulative number of observations of the observers collecting the raw data, surge is the mean reported surge across transects, kelp is the mean kelp coverage at the site, temp is the water temperature, regional_temp_dev is the water temperature at a given site scaled by the mean water temperature at that site. Targeted is a Boolean marker indicating whether the given observation is for targeted or non-targeted species.

@@

## Synthetic controls

@@

# Testing Model Assumptions

## Simulation testing

@@

## Testing SUTVA with Convergent Cross Mapping

@@

# Literature Cited

>>>>

---
title: "Supporting Information"
subtitle: "Assessing the Population-level Conservation Effects of Marine Protected Areas"
author: 
  - Daniel Ovando
  - Jennifer E. Caselle
  - Christopher Costello
  - Olivier Deschenes
  - Steven D. Gaines
  - Ray Hilborn
  - Owen Liu
output: 
  bookdown::pdf_document2:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: template.docx
  pdf_document: default
bibliography: references.yaml
csl: conservation-biology.csl
header-includes:
    - \usepackage{longtable}
    - \usepackage{booktabs}
    - \usepackage{eso-pic}
    - \usepackage{graphicx}
    - \usepackage[left]{lineno}
    - \usepackage{xcolor}
params:
  run_name: ["v6.0"]
  sim_years: [50]
  burn_years: [20]
  num_patches: [25]
always_allow_html: true
toc: false
---

@@

\renewcommand{\thesection}{Appendix S\arabic{section}}

@@

All data and code needed to fully reproduce replicate our study this manuscript are publicly available at from github.com/DanOvando/population-effects-of-mpas

All analysis were performed in `r R.version.string`. Package versions are shown in Table.\@ref(tab:pkg-info).

@@

```{r sim-table, echo=FALSE}
variable <- c(
"Scientific Name",
"steepness",
"Adult movement ($\\sigma_{s=a}$)",
"Larval movement ($\\sigma_{s=l}$)",
"Recruitment variation ($\\sigma_{r}$)",
"Recruitment autocorrelation ($ac_{r}$)",
"Density-dependent adult movement",
"Density-dependence timing ",
"% Patches in MPA",
"Initial fishing relative to natural mortality",
"Selectivity as a multiple of maturity length",
"Fleet model",
"Spatial effort model",
"Years into simulation to start MPA",
"MPA is Larval Source?",
"Randomly place MPA?",
"Fleet reaction to MPA",
"Patchiness",
"MPA habitat factor")

distribution <- 
c(
"Drawn from all possible species in `FishLife` (@thorson2017c)",
"uniform(0.6,0.95)",
"uniform(0,0.25 * P)",
"uniform(0,0.25 * P)",
 "$\\in{0,0.05,.1,.2}$",
 "$\\in{0,0.05,.1,.2}$",
 "$\\in{0.25,1}$",
 "$\\in{Local, Global,PostDispersal}$",
 "uniform(0.01,1)",
 "uniform(0.01,4)",
 "uniform(0.1,1.25)",
"$\\in{OpenAccess, ConstantEffort,ConstantCatch}$",
 "$\\in{Uniform, Biomass,Profits}$",
 "uniform(5,0.66)",
"$\\in{TRUE,FALSE}$",
"$\\in{TRUE,FALSE}$",
"$\\in{Concentrate,Leave}$",
"uniform(0.01,0.75)",
"$\\in{1, 4}$"
)

equations <- tibble(Variable = variable, Distribution = distribution )

knitr::kable(equations,
             escape = FALSE,
             caption = "Distributions of variables used in MPA simulations. Simulations combine random combinations of the variables included here, and using those sampled states simulate a scenario with and a scenario without MPAs. The difference between these two scenarios is then used to calcualte the effect of the MPA in that simulated scenario.",
             format = "pipe",
             align = "l") %>% 
  kableExtra::kable_styling()

@@

## Recruitment {-}

@@

## Dispersal {-}

@@

## Fleet Dynamics {-}

@@

## Spatial Fleet Distribution {-}

@@

## MPA Design {-}

@@

## Simulations{-}

@@

### Filtering Simulations {-}

@@

### Additional Simulation Results {-}

@@

The full table of covariates and their posterior means and 89% credible interval can be found in Table.\@ref(tab:coef-table) and Table.\@ref(tab:did-coefs). "var" refers to variable, "_2" indicates a squared term. tex is the mean cumulative number of observations of the observers collecting the raw data, surge is the mean reported surge across transects, kelp is the mean kelp coverage at the site, temp is the water temperature, regional_temp_dev is the water temperature at a given site scaled by the mean water temperature at that site. Targeted is a Boolean marker indicating whether the given observation is for targeted or non-targeted species.

```{r coef-table, echo=FALSE}

coefs <- did_fit %>%
  broom.mixed::tidy(conf.int = TRUE, prob = 0.9) %>%
    rename(lower = conf.low,
         upper = conf.high) %>% 
  select(term, estimate, lower, upper) %>%
  rename(
    Mean = estimate,
    `5th Percentile` = lower,
    `95th Percentile` = upper
  )

coefs$term <- str_remove_all(coefs$term, "var_")

coefs$term <- str_replace_all(coefs$term, "targeted","Targeted by Fishing")

coefs$term <- str_replace_all(coefs$term, "year_bins","Year Block ")

coefs$term <- str_replace_all(coefs$term, ":","$\\\\times$")

coefs <- coefs %>% 
  mutate(term = case_when(term == "tex" ~ "Observer Experience",
                          term == "tex_2" ~ "(Observer Experience)$^2$",
                          term == "temp" ~ "Water Temperature",
                          term == "regional_temp_dev" ~ "Deviation from regional mean temperature",
                          term == "regional_temp_dev_2" ~ "(Deviation from regional mean temperature)$^2$",
                          term == "lag_catch" ~ "Lagged Commercial Catch",
                          term == "targeted" ~ "Targeted by Fishing",
                          term == "(Intercept)" ~ "Intercept",
                          term == "surge" ~ "Wave Surge",
                          term == "kelp" ~ "Kelp Cover",
                           TRUE ~ term))  %>% 
  rename(Variable = term)

  knitr::kable(
    coefs,
    digits = 2,
    caption = glue::glue("Posterior means and credible interval for key model coefficients. Raw data contained {nrow(used_data)} individual observations. Total number of targeted and non-targeted mean total biomass densities by site and year used to fit model was {nrow(did_fit$data)}. Site-level terms omitted from table for visual clarity."),
    format = "pipe", 
    escape = FALSE,
    align = "l")

@@

## Synthetic controls {-}

@@

# Simulation Testing

@@

# Testing SUTVA with Convergent Cross Mapping

@@

# Appendix S8 Literature Cited

### 2021-05-02 documents/ovando-et-al-pop-effects-of-mpas.Rmd
pre-print cleanup

theme_1col <- hrbrthemes::theme_ipsum(base_size = 8,
                                      axis_title_size = 6,
                                      plot_title_size = 10,
                                      strip_text_size = 8)

theme_2col <- hrbrthemes::theme_ipsum(base_size = 10,
                                        axis_title_size = 14,
                                      plot_title_size = 12)

@@

All analysis were conducted in R [@rcoreteam2019]. Our DiD regression was fit using Stan [@carpenter2017] through the `rstanarm` package [@goodrich2020]. All data and code needed to fully reproduce this manuscript are publicly available at [github.com/DanOvando/population-effects-of-mpas](https://github.com/DanOvando/population-effects-of-mpas). A detailed description of the simulation model structure, as well as sensitivity analyses of our estimation model, are available in the online Appendix Supporting Information 1.

@@

Our bio-economic model simulates the effect of MPAs on a spatially explicit age-structured representation of a fish population. Readers can explore the functionality of the model using an online tool available at [danovando.shinyapps.io/simmpa/](https://danovando.shinyapps.io/simmpa/). The purpose of the simulation model is to both set expectations for our empirical results, and to demonstrate the ways in which ecological and economic dynamics can interact to produce a wide range of population-level MPA effects. The full range of factors explored are shown in  Table \@ref(tab:sim-table), and the equations of the simulation model can be viewed in the Appendix S1 Section 2. We used this model to generate `r n_distinct(outcomes$experiment)` simulated MPA outcomes across `r n_distinct(outcomes$scientific_name)` species.

@@

Fishing dynamics may be one factor contributing to a lack of strong MPA network effects. Much of the theoretical literature on MPAs assumes that larger reserves produce larger conservation gains (@white2011 and references therein). However, these models generally simulate fleet dynamics through fishing mortality rates; the proportion of total mortality experienced by a population attributable to fishing pressure (e.g. @halpern2004).  Alternatively, under a “constant-catch” strategy, fishers have a catch objective and exert as much (or little) effort as needed to achieve that objective. Subsistence fisheries may use a constant-catch style policy over the short-term, as they seek to ensure that their food needs are met. Constant-catch dynamics might also occur in fisheries with constraining quotas that are not updated after the implementation of MPAs. Fishers pursuing constant-catch in areas outside an MPA may have to fish harder to achieve the same catch from a smaller part of the population, causing a population conservation loss under `r percent(1 - positive_outcomes$pp[positive_outcomes$fleet_model == "constant-catch"], accuracy = 1)` of our constant-catch simulations. This potential negative interaction between constant-catch and MPAs is an important risk to consider (as done in @little2011b) especially as MPAs are increasingly implemented in quota-managed fisheries [@liu2018]. We do not have access to fine scale fishing data from the Channel Islands alone, but reported catches for the species of interest in the Santa Barbara region exhibit a mix of stable, downward, and upward trajectories (Appendix S1 Fig.S2) which indicates that a negative MPA effect caused by a constant-catch fishing strategy is unlikely.

@@

All of the species in this empirical analysis may affect each other through mechanisms such as predation, competition, and habitat modification. We used convergent cross mapping (CCM), in the manner of @clark2015, to test for significant dynamic interactions between species and therefore the possibility of the trophic cascades biasing our results. We found no significant cross-mappings between targeted and non-targeted species, indicating that while clearly there are interactions between these groups on some level, the effects within the timespan of the data are not pronounced enough to be of concern to our results (Appendix S1 Fig.S33-S35). However, the longer MPAs are in place, the greater the possibility that substantial species interactions that can affect use of non-targeted species as a control may arise.

@@

Detailed descriptions of the simulation model and DiD estimator (Appendix S1) are available online. The authors are solely responsible for the content and functionality of these materials. Queries (other than absence of the material) should be directed to the corresponding author.

@@

# Tables {#tables .unnumbered}

```{r sim-table, echo=FALSE}
variable <- c(
"Scientific Name",
"steepness",
"Adult movement ($\\sigma_{s=a}$)",
"Larval movement ($\\sigma_{s=l}$)",
"Recruitment variation ($\\sigma_{r}$)",
"Recruitment autocorrelation ($ac_{r}$)",
"Density-dependent adult movement",
"Density-dependence timing ",
"% Patches in MPA",
"Initial fishing relative to natural mortality",
"Selectivity as a multiple of maturity length",
"Fleet model",
"Spatial effort model",
"Years into simulation to start MPA",
"MPA is Larval Source?",
"Randomly place MPA?",
"Fleet reaction to MPA",
"Patchiness",
"MPA habitat factor")

distribution <- 
c(
"Drawn from all possible species in `FishLife` (@thorson2017c)",
"uniform(0.6,0.95)",
"uniform(0,0.25 * P)",
"uniform(0,0.25 * P)",
 "$\\in{0,0.05,.1,.2}$",
 "$\\in{0,0.05,.1,.2}$",
 "$\\in{0.25,1}$",
 "$\\in{Local, Global,PostDispersal}$",
 "uniform(0.01,1)",
 "uniform(0.01,4)",
 "uniform(0.1,1.25)",
"$\\in{OpenAccess, ConstantEffort,ConstantCatch}$",
 "$\\in{Uniform, Biomass,Profits}$",
 "uniform(5,0.66)",
"$\\in{TRUE,FALSE}$",
"$\\in{TRUE,FALSE}$",
"$\\in{Concentrate,Leave}$",
"uniform(0.01,0.75)",
"$\\in{1, 4}$"
)

equations <- tibble(Variable = variable, Distribution = distribution )

knitr::kable(equations,
             escape = FALSE,
             caption = "Distributions of variables for MPA simulations",
             format = "pipe",
             align = "l") %>% 
  kableExtra::kable_styling()

```{r coef-table, echo=FALSE}

coefs <- did_fit %>%
  broom.mixed::tidy(conf.int = TRUE, prob = 0.9) %>%
    rename(lower = conf.low,
         upper = conf.high) %>% 
  select(term, estimate, lower, upper) %>%
  rename(
    Mean = estimate,
    `5th Percentile` = lower,
    `95th Percentile` = upper
  )

coefs$term <- str_remove_all(coefs$term, "var_")

coefs$term <- str_replace_all(coefs$term, "targeted","Targeted by Fishing")

coefs$term <- str_replace_all(coefs$term, "year_bins","Year Block ")

coefs$term <- str_replace_all(coefs$term, ":","$\\\\times$")

coefs <- coefs %>% 
  mutate(term = case_when(term == "tex" ~ "Observer Experience",
                          term == "tex_2" ~ "(Observer Experience)$^2$",
                          term == "temp" ~ "Water Temperature",
                          term == "regional_temp_dev" ~ "Deviation from regional mean temperature",
                          term == "regional_temp_dev_2" ~ "(Deviation from regional mean temperature)$^2$",
                          term == "lag_catch" ~ "Lagged Commercial Catch",
                          term == "targeted" ~ "Targeted by Fishing",
                          term == "(Intercept)" ~ "Intercept",
                          term == "surge" ~ "Wave Surge",
                          term == "kelp" ~ "Kelp Cover",
                           TRUE ~ term))  %>% 
  rename(Variable = term)

  knitr::kable(
    coefs,
    digits = 2,
    caption = glue::glue("Posterior means and credible interval for key model coefficients. Raw data contained {nrow(used_data)} individual observations. Total number of targeted and non-targeted mean total biomass densities by site and year used to fit model was {nrow(did_fit$data)}. Site-level terms omitted from table for visual clarity."),
    format = "pipe", 
    escape = FALSE,
    align = "l")

@@

temp_plot <- response_ratio_plot + 
  theme_1col + 
  theme(axis.title.x = element_textbox(size = 10),
        legend.position = "top",
        panel.spacing.y = unit(1, "lines"))

@@

ggsave(
  mpa_effect_plot,
  filename = here::here("documents","figs","mpa_effect_plot.pdf"),
  width = width_2col,
  height = width_2col * asp_2col,
  device = cairo_pdf
)

>>>>

theme_1col <- hrbrthemes::theme_ipsum(base_size = 8,
                                      axis_title_size = 6,
                                      plot_title_size = 10,
                                      strip_text_size = 8) + 
    theme(axis.title.x = element_text(hjust = 0.5),
        axis.title.y = element_text(hjust = 0.5))

theme_2col <- hrbrthemes::theme_ipsum(base_size = 10,
                                        axis_title_size = 14,
                                      plot_title_size = 12) + 
    theme(axis.title.x = element_text(hjust = 0.5),
        axis.title.y = element_text(hjust = 0.5))

@@

All analysis were conducted in R [@rcoreteam2019]. Our DiD regression was fit using Stan [@carpenter2017] through the `rstanarm` package [@goodrich2020]. All data and code needed to fully reproduce this manuscript are publicly available at [github.com/DanOvando/population-effects-of-mpas](https://github.com/DanOvando/population-effects-of-mpas). A detailed description of the simulation model structure, as well as sensitivity analyses of our estimation model, are available in the online Appendix S2.

@@

Our bio-economic model simulates the effect of MPAs on a spatially explicit age-structured representation of a fish population. Readers can explore the functionality of the model using an online tool available at [danovando.shinyapps.io/simmpa/](https://danovando.shinyapps.io/simmpa/). The purpose of the simulation model is to both set expectations for our empirical results, and to demonstrate the ways in which ecological and economic dynamics can interact to produce a wide range of population-level MPA effects. The full range of factors explored are shown in  Table \@ref(tab:sim-table), and the equations of the simulation model can be viewed in the Appendix S2. We used this model to generate `r n_distinct(outcomes$experiment)` simulated MPA outcomes across `r n_distinct(outcomes$scientific_name)` species.

@@

Fishing dynamics may be one factor contributing to a lack of strong MPA network effects. Much of the theoretical literature on MPAs assumes that larger reserves produce larger conservation gains (@white2011 and references therein). However, these models generally simulate fleet dynamics through fishing mortality rates; the proportion of total mortality experienced by a population attributable to fishing pressure (e.g. @halpern2004).  Alternatively, under a “constant-catch” strategy, fishers have a catch objective and exert as much (or little) effort as needed to achieve that objective. Subsistence fisheries may use a constant-catch style policy over the short-term, as they seek to ensure that their food needs are met. Constant-catch dynamics might also occur in fisheries with constraining quotas that are not updated after the implementation of MPAs. Fishers pursuing constant-catch in areas outside an MPA may have to fish harder to achieve the same catch from a smaller part of the population, causing a population conservation loss under `r percent(1 - positive_outcomes$pp[positive_outcomes$fleet_model == "constant-catch"], accuracy = 1)` of our constant-catch simulations. This potential negative interaction between constant-catch and MPAs is an important risk to consider (as done in @little2011b) especially as MPAs are increasingly implemented in quota-managed fisheries [@liu2018]. We do not have access to fine scale fishing data from the Channel Islands alone, but reported catches for the species of interest in the Santa Barbara region exhibit a mix of stable, downward, and upward trajectories (Appendix S3) which indicates that a negative MPA effect caused by a constant-catch fishing strategy is unlikely.

@@

All of the species in this empirical analysis may affect each other through mechanisms such as predation, competition, and habitat modification. We used convergent cross mapping (CCM), in the manner of @clark2015, to test for significant dynamic interactions between species and therefore the possibility of the trophic cascades biasing our results. We found no significant cross-mappings between targeted and non-targeted species, indicating that while clearly there are interactions between these groups on some level, the effects within the timespan of the data are not pronounced enough to be of concern to our results (Appendix S7). However, the longer MPAs are in place, the greater the possibility that substantial species interactions that can affect use of non-targeted species as a control may arise.

@@

Detailed descriptions of the simulation model and DiD estimator are available in the online supporting information The authors are solely responsible for the content and functionality of these materials. Queries (other than absence of the material) should be directed to the corresponding author.

@@

@@

temp_plot <- response_ratio_plot + 
  theme_1col + 
  theme(
        legend.position = "top",
        panel.spacing.y = unit(1, "lines"))

@@

ggsave(
  mpa_effect_plot + theme_2col + theme(legend.position = "top"),
  filename = here::here("documents","figs","mpa_effect_plot.pdf"),
  width = width_2col,
  height = width_2col * asp_2col,
  device = cairo_pdf
)

### 2021-05-02 documents/pop-effects-of-mpas-front-matter.Rmd
pre-print cleanup

Jennifer E. Caselle
UCSB Marine Science Institute Bldg 520 Rm 4002 Fl 4L 
Santa Barbara, CA 93106-6150 USA
caselle@msi.ucsb.edu

Christopher Costello
Bren Hall 4410 
University of California, Santa Barbara
Santa Barbara, CA 93106-6150 USA
costello@bren.ucsb.edu

Olivier Deschenes
2127 North Hall
University of California, Santa Barbara
Santa Barbara, CA 93106-6150 USA
olivier@econ.ucsb.edu

Steven D. Gaines
Bren Hall 2410 
University of California, Santa Barbara
Santa Barbara, CA 93106-6150 USA
gaines@bren.ucsb.edu

Ray Hilborn
FSH 352B
University of Washington
1122 NE Boat Street
Seattle, WA 98195-5020 USA

Owen Liu
Bren Hall 4330 
University of California, Santa Barbara
Santa Barbara, CA 93106-6150 United States
owen.r.liu@gmail.com

@@

Funding for the simulation model was provided by the National Marine Fisheries Service-SeaGrant Population and Ecosystem Dynamics Fellowship. Empirical data collection was funded primarily by the David and Lucille Packard Foundation in support of the Partnership for Interdisciplinary Studies of Coastal Oceans (PISCO) with additional funding from the California Ocean Protection Council and California SeaGrant. This study would not have been possible without the work provided by PISCO divers over the years, with special thanks to K. Davis Koehn and A. Parsons-Field. We also thank C. Szuwalski, J. Lawson, A. Punt, C. Brown, and two anonymous reviewers for helpful comments and technical support.

>>>>

Daniel Ovando*
School of Aquatic and Fishery Sciences
University of Washington, Seattle, WA 98195-5020 USA
danovan@uw.edu

Jennifer E. Caselle
Marine Science Institute, University of California Santa Barbara, Santa Barbara, CA, USA

Christopher Costello
Environmental Market Solutions Lab, University of California, Santa Barbara, CA, USA

Olivier Deschenes
Environmental Market Solutions Lab, University of California, Santa Barbara, CA, USA 
Department of Economics, University of California, Santa Barbara, CA, USA

Steven D. Gaines
Bren School of Environmental Science and Management, University of California, Santa Barbara, CA 93117

Ray Hilborn
School of Aquatic and Fishery Sciences
University of Washington, Seattle, WA 98195-5020 USA

Owen Liu
Bren School of Environmental Science and Management, University of California, Santa Barbara, CA 93117

@@

Funding for the simulation model was provided by the National Marine Fisheries Service-SeaGrant Population and Ecosystem Dynamics Fellowship. Empirical data collection was funded primarily by the David and Lucille Packard Foundation in support of the Partnership for Interdisciplinary Studies of Coastal Oceans (PISCO) with additional funding from the California Ocean Protection Council and California SeaGrant. This study would not have been possible without the work provided by PISCO divers over the years. We especially thank K. Davis Koehn, and A. Parsons-Field. We also thank C. Szuwalski, J. Lawson, A. Punt, C. Brown, and 2 anonymous reviewers for helpful comments and technical support.
